sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-16 doc/paper.Rmd
add results from appendix 4 and 5 to paper

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
<!-- TODO: still true? -->
There is considerable variability in the available observation likelihoods across packages (Table 1).

@@

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing)).
In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern substantially (not shown).
Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary difference is the user interface.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA may find inlabru approachable (Appendix 4).
Users already familiar with mgcv can fit adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2021].
INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.

>>>>

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
<!-- TODO: still true? -->
There is considerable variability in the available observation likelihoods across packages (Table 1).
<!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
We provide examples of the syntax and the reproducibility of results achieved when similar models were built using INLA and sdmTMB (Appendix 4) and VAST and sdmTMB (Appendix 5). 
We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax. <!-- Change this if we manage to solve the inlabru issue  -->

@@

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Appendix 6 TODO, Fig. \@ref(fig:timing)).
In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern substantially (not shown).
Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary difference is the user interface.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA may find inlabru approachable. <!-- ; however, we ran into problems with AR1 spatiotemporal fields in inlabru, so we only provide a comparison with INLA instead (Appendix 4). -->
Users already familiar with mgcv can fit adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2021].
INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.
